A multi-scale story of the diffusion of a new technology: the web

### 2024-08-14 paper/submission1/tranos2023.tex
postpresentation

\section{Introduction}\label{sec1}

@@

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related \emph{future} technologies. As per
\citet{leibowicz2016representing}, historical studies agree on the fact
that ``technologies diffuse at different times, at different rates, and
to different extents in different places, and can be significantly
influenced by policies'' \citep{victor1993}. \citet{meade2021modelling}
highlight that a variety of actors have a direct interest in gaining
such knowledge including network equipment suppliers; network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places \citep{leibowicz2016representing}. Despite the spatial
heterogeneity of such diffusion mechanisms and the policy relevance,
there are very limited attempts in the literature to analyse the
diffusion of new digital technologies at a detailed geographical level.

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, \citet{rogers2010diffusion} identifies early adopter of new
technologies as `knowledgeable risk takers' and \citet{griliches1957} as
`profit maximisers' \citep{ding2010modeling}. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, \citet{perkins2011internet} attribute
diffusion to two processes: (i) epidemic-like mechanisms, which are
governed by distance, proximity and social interactions, and (ii) by
economic mechanisms as new innovations are adopted by users as they
become more profitable, valueable and useful.

@@

Importantly, the data used here depicts the \emph{active} engagement
with the web in the UK as it contains all commercial websites that (i)
are part of the UK's relevant second level domain (SLD, .co.uk), (ii)
have been archived by the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.},
and (iii) include a mention to at least one valid UK postcode in the web
text. The act of creating a website is understood here as active
engagement with the web vis-à-vis the more passive act of browsing the
web or having an internet connection \citep{tranosuk}. Previous studies
have focused mostly on more passive notions of engaging with digital
technologies such as internet adoption and internet speeds
\citep[e.g.][]{blank2018local, riddlesden2014broadband}. More details
about the data and the data generation process can be found in Section
\hyperref[sec3]{XX}.

@@

\section{Literature review}\label{sec2}

Geographical diffusion is a synthesis of different processes. On the one
hand, we can identify purely spatial or, in other words, contagious
processes. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, we can identify hierarchical processes. Instead of
horizontal distance-based diffusion mechanisms, the top-down hierarchy
of urban systems drives diffusion. In reality, the synthesis of these
two processes represents how new technologies diffuse over space and
time \citep{morrill2020spatial}.

@@

The temporal dimension was further explored by Everett Rogers and his
seminal work on `Diffusion of Innovations' \citep{rogers2010diffusion}.
Rogers being a sociologist, his work focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism
of diffusion and adoption is communication and how knowledge is
transferred within a social system. Therefore, all approaches agree on
the S-shaped diffusion and cumulative adoption pattern
\citep{grubler1990rise}.

Schmidt's Law empirically illustrates a similar pattern. \emph{Core} and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed \citep{grubler1990rise}. This is
where the first adopters tend to be based. Then, technologies spread to
the \emph{rim} and eventually to the \emph{periphery}. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions \citep{leibowicz2016representing}. \citet{grubler1990rise}
effectively summarises the three key characteristics of the spatial
diffusion process: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchy effect in a form of a centrifugal force: from core
to periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the
three mechanisms that the empirical analysis in Section
\phantomsection\label{sec4}{4} investigates.

@@

\section{Materials and Methods}\label{sec3}

@@

To capture the diffusion of web technologies, a website density metric
is developed for two different geographical scales: the Local Authority
Districts (LAD) and the Output Areas (OA). The former is an
administrative unit and there are
\textbackslash textcolor\{red\}c.~380\textbackslash textcolor\{black\}
such units in the UK. The latter is a census-based geographical unit,
which is very small as there are c.~230,000 of them in the UK. This
methodological choice will allow the mapping of the diffusion of web
technologies and the assessment of the diffusion mechanisms at these two
very different spatial scales.

The population of websites is calculated using data from the Internet
Archive \footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage its
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK ccTLD (.uk) from the
1996--2012 period. In essence, this is a long list of 2.5 billion URLs
of archived webpages including also the archival timestamp.

Instead of using the whole .uk country code top-level domain (ccTLD),
this paper focuses on its commercial subset, the .co.uk SLD. This choice
decrease the heterogeneity of the web data as such commercial websites
have rather specific aims: they are used to diffuse information, support
online transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as.com and such cases escape the data used
here, such omissions should not affect our results given the popularity
of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to visit a .uk
website when they are searching for products or services \citep{hope};
and anecdotal evidence indicates that during the first half of 2000,
three .co.uk domains were registered every minute \citep{oecd_coms}.
Importantly, previous studies illustrated that .co.uk is the most
popular UK SLD \citep{tranosuk}.

The text from these webpages was scanned using regular expressions
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts.

Before that, the data cleaning process included a data aggregation step,
through which the webpages were aggregated to the relevant websites.
Based on the following example, all three webpages listed below are part
of the same website (\url{http://www.website.co.uk}) and at the end only
websites and not the nested webpages were considered as otherwise the
metric would have been biased towards large websites.

\begin{itemize}
\item
  \url{http://www.website.co.uk/webpage_a} B152TT
\item
  \url{http://www.website.co.uk/webpage_b} BS81TH
\item
  \url{http://www.website.co.uk/webpage_c} B152TT
\end{itemize}

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
(\url{http://www.website.co.uk}) includes two unique postcodes: B152TT
and BS81SS. The distribution of postcodes for 2000 is presented in Table
\ref{f2000}, which clearly illustrates the wide range. On one hand, we
have websites anchored to a unique location (72\% of all the
reconstructed websites), which may represent a small company with a
single trading location. On the other hand, we have websites with
thousands of different postcodes. Considering the timeperiod of the
analysis, such cases can represent directories which used to be popular
in the pre-search engines early times of internet
\citep{tranosuk}.\footnote{See Figure A1 in \textbf{Appendix A in the
  supplemental data online for examples}.}

The analysis presented here is based on two subsets of these data.
Firstly, the analysis is based on websites which only contain only one
unique postcode and, therefore, the geolocation process might suffer
from less noise. As a robustness check, the analysis is replicated the
analysis for an extended subset of websites, which include up to ten
unique postcodes. These websites are geolocated by equally attaching
them to multiple locations. This extended sample includes 94 percent of
all the archived websites in 2000.

@@

To create the web density metric, the yearly website counts at the LAD
level are standardised by the number of firms in LAD to avoid biases
associated with LADs hosting a large number of firms. Given that there
is no such statistic for the OA, the actual OA level counts are used.
However, because of the the rather consistent spatial definition of OA
(they host 40-250 households),\footnote{\url{https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies}}
website counts in OA are interpreted as a density metric too.

This website metric\ldots{}

\citet{wilson201281} growth function description in p.~86

from R: Asym/(1+exp((xmid-input)/scal)) using the terms from Wilson k/(1
+ e\^{}(t\_0 - t)/scal)

Wilson: k(1 + e\^{}-b(t - t\_0)) So, scal = - 1/b

The curve is symmetric* around t\_0

\textbf{minimum R2 = 95\%}, see also \citet{grubler1990rise}

The literature usually uses the saturation level as the asymptote. I am
using the total number of websites as we cannot compute a rate.

Moran's I as \citet{ding2010modeling} \textbf{TODO} for t\_0, diffusion
speed

\section{Results}\label{sec4}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  S-shaped diffusion curves: S for LADs per firm. UK, fast/slow/examples
\item
  ranks: there is stability and movement
\item
  Neighbourhood effect: diffusion proceeds outwards from innovation
  centers, first ``hitting'' nearby rather than far-away locations
  (Grubler 1990)
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  ideal: (i) train RF for all years and all (1) LADs and (2) OAs with
  CAST and report metrics. (ii) train for all years and all but one
  region for (1) LADs and (2) OAs to predict to the holdout region.
  Reports predictions as region similarities.
\end{itemize}

\subsubsection{RF results}\label{rf-results}

The next section incorporates the above discussed spatial processes of
the diffusion of web technologies into the same modelling framework. The
aim is to use variables depicting these spatial processes in order to
predict the diffusion of web technologies in the UK over space and time
and across different scales. Specifically, four different models are
estimated. Firstly, all the data points at the OA and LAD are utilised
in order to build RF models and assess their capacity to predict the
adoption of web technologies. These two models will reveal the
predictive capacity of the spatial processes and also allow to see how
the importance of such variables changes across scales. The next two
sets of models will be again trained on web diffusion at the two working
scales: OA and LAD. However, instead of using all the data points, the
OA and the LAD from one of the twelve UK regions are held out and then
the trained model is used to predict web adoption in the OA or the LAD
of the held-out region. This process takes place recursively for all UK
regions. The difference in the predictive capacity of the different
samples will reveal how dissimilar are these spatial process across
regions and, importantly, at different scales.

@@

The models need to include variables that capture the three processes
that the relevant literature and the descriptive analysis presented in
the Section \textbf{XX} highlighted. Namely, the models capture: (i) a
hierarchy effect with diffusion running from main centres to secondary
ones, (ii) a neighborhood effect according to which diffusion first hits
nearby locations, and (iii) the rather canonical pattern of diffusion
over time as reflected in the S-shaped pattern in the cumulative level
of adoption.

>>>>

\section{Introduction}\label{sec-introduction}

@@

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related \emph{future} technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers; network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there are very limited attempts in the literature to
analyse the diffusion of new digital technologies at a detailed
geographical level.

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, \citet{rogers2010diffusion} identifies early adopter of new
technologies as `knowledgeable risk takers' and \citet{griliches1957} as
`profit maximisers' \citep{ding2010modeling}. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, \citet{perkins2011internet} attribute
diffusion to two processes: (i) epidemic-like mechanisms, which are
governed by distance, proximity and social interactions, and (ii)
economic mechanisms as new innovations are adopted by users as they
become more profitable, valueable and useful.

@@

Importantly, the data used here depicts the \emph{active} engagement
with the web in the UK as it contains all commercial websites that (i)
are part of the UK's relevant second level domain (SLD, .co.uk), (ii)
have been archived by the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.},
and (iii) include a mention to at least one valid UK postcode in the web
text. The act of creating a website is understood here as active
engagement with the web vis-à-vis the more passive act of browsing the
web or having an internet connection \citep{tranosuk}. Previous studies
have focused mostly on more passive notions of engaging with digital
technologies such as internet adoption and internet speeds
\citep[e.g.][]{blank2018local, riddlesden2014broadband}. More details
about the data and the data generation process can be found in
Section~\ref{sec-datamethods}.

\textbf{S curve??}

@@

\section{Literature review}\label{sec-litreview}

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be
identified. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure \citep{hivner2003facilitating}. On
the other hand, there is a hierarchical processes. Instead of horizontal
distance-based diffusion mechanisms, the top-down hierarchy of urban
systems shapes technological diffusion. In reality, the synthesis of
these two processes represents how new technologies diffuse over space
and time \citep{morrill2020spatial}.

@@

The temporal dimension was further explored by Everett Rogers and his
seminal work on `Diffusion of Innovations' \citep{rogers2010diffusion}.
Rogers being a sociologist focused not on the diffusion of innovations
over space and time, but instead on the adoption of new technologies and
innovations by individuals and the individual mechanisms that drive the
decisions behind adoption. He identified five groups of individuals
regarding their adoption speed: innovators, early adopters, early
majority, late majority and laggards. The key mechanism of diffusion and
adoption is communication and how knowledge is transferred within a
social system. Therefore, all approaches agree on the S-shaped diffusion
and cumulative adoption pattern \citep{grubler1990rise}.

Schmidt's Law empirically illustrates a similar pattern. \emph{Core} and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed \citep{grubler1990rise}. This is
where the first adopters tend to be based. Then, technologies spread to
the \emph{rim} and eventually to the \emph{periphery}. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions \citep{leibowicz2016representing}. \citet{grubler1990rise}
effectively summarises the three key characteristics of the spatial
diffusion process: (i) the cumulative level of adoption follows an
S-shaped pattern just like purely temporal models; (ii) diffusion is
shaped by a hierarchy effect in a form of a centrifugal force: from core
to periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the
three mechanisms that the empirical analysis in
Section~\ref{sec-results} investigates.

@@

\section{Data and Methods}\label{sec-datamethods}

@@

To capture the diffusion of web technologies, a website density metric
is developed for two different geographical scales: the Local Authority
Districts (LAD) and the Output Areas (OA). The former is an
administrative unit and there are c.~374 such units in the UK. The
latter is a census-based geographical unit, which is very small as there
are c.~230,000 of them in the UK. This methodological choice will allow
the mapping of the diffusion of web technologies and the assessment of
the diffusion mechanisms at these two very different spatial scales.

The counts of websites at these scales are calculated using data from
the Internet Archive \footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage its
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK ccTLD (.uk) from the
1996--2012 period. In essence, this is a long list of 2.5 billion URLs
of archived webpages including also the archival timestamp.

Instead of using the whole .uk country code top-level domain (ccTLD),
this paper focuses on its commercial subset, the .co.uk SLD. This choice
decrease the heterogeneity of the web data as such commercial websites
have rather specific aims: they are used to diffuse information, support
online transactions and share opinions
\citep{THELWALL2000441, blazquez2018big}. Although a UK company can
adopt a generic TLD such as .com and these cases escape the data used
here, such omissions should not affect our results given the popularity
of the .uk ccTLD \citep{tranosuk}: UK consumers prefer to visit a .uk
website when they are searching for products or services \citep{hope};
and anecdotal evidence indicates that during the first half of 2000,
three .co.uk domains were registered every minute \citep{oecd_coms}.
Importantly, previous studies illustrated that .co.uk is the most
popular UK SLD \citep{tranosuk}.

The text from these webpages was scanned using a regular expression
(regex) to identify strings of text which resemble UK postcodes and one
fifth of them included a mention to a postcode \citep{BL2013geo}. This
information allows the geolocation of the data and the creation of the
LAD and OA counts.

The data cleaning process included an aggregation step, through which
the archived webpages were aggregated to the parent websites. This
website reconstruction allows the creation of a \emph{website} instead
of a \emph{webpage} density metric. Websites, which are hierarchical
structures, tend to represent specific orgnanisations or entities, and,
arguably, are more meaningful observational units than webpages, which
ignore the upstream dependencies. Based on the following example, all
three webpages are part of the same website
(\url{http://www.website.co.uk}) and at the end only websites and not
the nested webpages were considered as otherwise the metric would have
been biased towards large, place specific, websites.

\begin{itemize}
\item
  \url{http://www.website.co.uk/webpage_a} B15 2TT
\item
  \url{http://www.website.co.uk/webpage_b} BS8 1TH
\item
  \url{http://www.website.co.uk/webpage_c} B15 2TT
\end{itemize}

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
\href{http://www.website.co.uk}{www.website.co.uk} includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of postcodes for 2000
is presented in Table \ref{f2000}, which clearly illustrates the wide
range. At the left end of the distribution, there are websites anchored
to a unique location (72\% of all the reconstructed websites in 2000),
which may represent a small company with a single trading location. At
the right end, we have websites with thousands of different postcodes.
Considering the time period of the analysis, such cases can represent
directories which used to be popular in the pre-search engines early
times of the commercial internet \citep{tranosuk}.\footnote{See Figure
  A1 in \textbf{Appendix A in the supplemental data online for
  examples}.}

The analysis presented here is based on two subsets of these data.
Firstly, on websites, which only contain only one unique postcode and,
therefore, the geolocation process does not suffer from noise. As a
robustness check, the analysis is replicated for an extended subset of
websites, which include up to ten unique postcodes. These websites are
geolocated by equally attaching them to multiple locations. This
extended sample includes 94 percent of all the archived websites in
2000.

@@

Another data cleaning step dealt with some extreme outliers. Figure
\ref{pc1} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2004-2006 period, which can be
attributed to a link farm \citep{BL2013links}.\\
The website counts for these six postcodes (SE24 9HP, CV8 2ED, GL16 7YA,
CW1 6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. As the Section~\ref{sec-results} illustrates, this led
to a small increase of the model predictive capability. To put the
magnitude of the data imputation into perspective, this process affected
6 out of the 557,808 postcodes present in the data.

{\centering \includegraphics[width=1\textwidth,height=0.6\textheight]{../../outputs/pc_year_1.png}

\caption{\label{pc1}Yearly website counts per postcode (top 1000)}

To create the website density metric, the yearly website counts at the
LAD level are standardised by the number of firms in LAD to avoid biases
associated with LAD hosting a large number of firms. Given that there is
no such statistic for the OA, the actual OA level counts are used.
However, because of the the consistent spatial definition of OA (they
host 40-250 households),\footnote{\url{https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies}}
website counts in OA are interpreted as a density metric too.

These website density metrics are used for the three different steps of
the analysis. Firstly, a system-level analysis explores how the
diffusion of web technologies in the UK fits with the well-established
S-curve. To do so, the following logistic function (Equation \ref{eq:s})
is estimated for the whole of the UK and for each LAD separately:

\begin{align}
y = k /(1 + e^{-(t-t_{o})})\label{eq:s}
\end{align}

\noindent \(k\) is the asymptote or, in other words, the saturation
level, \(b\) the overall growth rate, and \(t_{0}\) the \emph{inflection
point} of maximum growth at \(k/2\), where the logistic function is
symmetrical \citep{wilson201281}. We compare the LAD \(t_0\) of each LAD
against the \(t_0\) of the UK to delineate whether a LAD reached that
point faster or slower than the country average. Importantly, an
accuracy criterion was imposed and only S-curves with \(R^2 > 0.9\) were
included in the analysis. To estimate Equation \ref{eq:s} a
self-starting logistic growth model was employed using the \texttt{nls}
and \texttt{SSlogis} functions in \texttt{R}.

The system-level analysis also focuses on the volatility of the web
adoption to depict places with high concentration of early adopters and,
equally, latecomers. To do so, the change of the ranking of the UK LAD
over time is plotted and discussed. Both the S-curve and the volatility
analysis focus only on LAD as the very large number of OA would have
made such analysis diffucult to visualise and interpret.

Next, exploratory spatial data analysis depicts whether the two main
drivers of spatial diffusion -- namely neighbourhood and hierarchy --
underpin the diffusion of web technologies in the UK. To capture the
former and following \citet{ding2010modeling} the Moran's I and the
Local Index of Spatial Association (LISA) are estimated for website
density. To address the hierarchy mechanism, the Gini coefficient -- a
well established metric of inequality -- is calculated. All the above
are computed and plotted longitudinally both for the LAD and the OA.

Lastly, a modelling framework is developed to test the above diffusion
mechanisms. The overarching aim is to build a model that can test the
relationship between the website density and these mechanisms:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S-curve_{t}\label{eq:rf.generic}
\end{align}

To estimate Equation \ref{eq:rf.generic}, Random Forest (RF) models were
built. This is a popular machine learning algorith for both regression
and classification problems \citep{biau2012analysis}. It was introduced
by \citet{breiman2001random} and has\\
gained popularity, becoming a go-to data science tool. RF can
effectively handle skewed distributions and outliers, model non-linear
relationships, require minimal hyperparameter tuning, exhibit low
sensitivity to these parameters, and have relatively short training
times \citep{Caruana2008, liaw2002classification, yan2020using}. These
attributes match well with the website density data characteristics
including skewness especially for OA. Also, the large data size (c.~230k
data points for each of the 17 years) call for fast training times.
Importantly, RF predictions tend to be more accurate than those from
single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors
\citep{mullainathan2017machine, athey2019machine, sulaiman2011intelligent, pourebrahim2019trip, biau2012analysis}.

RF is a tree-based ensemble learning algorithm
\citep{breiman2001random}. It begins by generating random samples of the
training data, which are then used to grow regression trees to predict
the dependent variable. Data points and predictors are randomly sampled
for the different trees. The trees are trained in parallel using their
own bootstrapped samples of the training data. A crucial feature of RF
is their ability not to overfit, meaning they can generalize well to
unseen test data. While each tree may overfit individually, the ensemble
of trees does not because the errors of individual trees are averaged,
reducing the overall variance and preventing overfitting
\citep{last2002improving}. For regression problems, RF predictions are
made by averaging the predictions of all decision trees.

RF have been widely employed to address regression research problems.
\citet{pourebrahim2019trip} combined a spatial interaction modeling
framework with ML algorithms including RF to predict commuting flows in
New York City. \citet{sinha2019assessing} advocated for adopting spatial
ensemble learning approaches, such as RF, to model spatial data with
high autocorrelation and heterogeneity. \citet{creditspatial} predicted
employment density in Los Angeles using spatially explicit RF.
\citet{guns2014recommending} used RF to build a recommendation system
for research collaborations. \citet{ren2019predicting} trained RF to
predict the socio-economic status of cities using various online and
mobility predictors. \citet{tranos2023using} utilised hyperlinks data
and RF to make out-of-sample predictions of interregional trade.
\citet{zhou2023geography} employed such a framework to assess whether
key predictors of obesity differ across English cities.

\section{Results}\label{sec-results}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  S-shaped diffusion curves: S for LAD per firm. UK, fast/slow/examples
\item
  ranks: there is stability and movement
\item
  Neighbourhood effect: diffusion proceeds outwards from innovation
  centers, first ``hitting'' nearby rather than far-away locations
  (Grubler 1990)
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  ideal: (i) train RF for all years and all (1) LAD and (2) OA with CAST
  and report metrics. (ii) train for all years and all but one region
  for (1) LAD and (2) OA to predict to the holdout region. Reports
  predictions as region similarities.
\end{itemize}

Figure \ref{s_uk} plots the S curve for the cumulative adoption of
website technologies in the UK during the 1996-2012 period. It
demonstrates a pattern well aligned with previous studies discussed in
Section~\ref{sec-litreview}. The vertical line for year 2003 illustrates
the point where the modeled cumulative adoption was equal to 50\% of the
maximum. This \(t_0\) \emph{inflection} point signals the maximum
adoption speed and is used here to determine whether the UK LAD reached
that inflection point earlier or later than the UK average.
Specifically, the S curve and the inflection point are estimated for
every LAD individually and then compared with the UK average. LAD that
reached their inflection point earlier than the country average are
labelled as \emph{fast} and the rest as \emph{slow}. Figure \ref{s_map}
maps this pattern and the picture is not entirely clear. On the one
hand, there are some expected examples of LAD with relevant industrial
backgrounds delineated as fast: the City of London, a world-renowned
cluster of finance industries \citep{cook2007role}, and Reading, a town
with high-tech service industries in proximity to London and its main
airport, Heathrow \citep{england2005polynet}. On the other hand, LAD
which were expected to appear as fast -- e.g.~Hackney in central London
and Bristol, a well-established creative cluster
\citep{oatley1999cultural, bassett2002cultural} -- were delineated as
slow. Nevertheless, out of the 10 fastest LAD, eight were located in
Greater London and one in Cambridge (see \textbf{Appendix}), but in
overall there is a spatially heterogenous and not easy to explain
spatial pattern.

{\centering \includegraphics[width=1\textwidth,height=\textheight]{../../outputs/s/s_uk_per_firm.png}

\caption{\label{s_uk}Grwoth curve, UK}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{../../outputs/s/speed_map.png}

\caption{\label{s_map}LAD adoption rates}

The next step is to assess the stability and volatility of the LAD in
terms of their adoption of web technologies. As we know from the
literature \citep{risk_perceptions}, different agents have different
perceptions about and levels of acceptance of the risks and the
potential economic returns associated with the adoption of new
technologies -- see for instance the seminal work of
\citet{venkatesh2000theoretical}. To reveal such aggregated patterns,
Figures \ref{rank_stable} and \ref{rank_unstable} plot the ranking of UK
LAD based on website density. To decrease noise, the average ranking of
1996-1998 and 2010-2012 is plotted instead of the individual years.
While there are quite a few obvious cases of LAD that maintained their
position between the beginning and the end of the study period either at
the top or at the bottom of the hierarchy (Figure \ref{rank_stable}),
there are also quite a few LAD that changed drastically their position.
Some of these LAD enjoyed a process that at the first instance looks
like leapfrogging since they managed to jump at the top of the hierarchy
despite their slow start (top of Figure \ref{unstable}). There is
extensive literature regarding the potential benefits of technological
leapfrogging. The underpinning argument is that latecomers can adopt and
benefit from new technologies that have been developed elsewhere without
incurring the hefty initial R\&D costs \citep{teece2008firm}. Although
the leapfrogging literature does not pay much attention to cities and
regions \citep{yu2018sustainability}, previous research highlighted the
long term and sustained productivity benefit of the early adopters of
web technologies in the UK \citep{tranosuk}. So, although the LAD system
is volatile, it is not clear whether the LAD with high concentration of
late adopters will gain any latecomer benefits in a way similar to
countries experiencing such technological leapfrogging.

{\centering \includegraphics[width=1\textwidth,height=\textheight]{../../outputs/ranks/web_per_firm2000_2012_only05_av.png}

\caption{\label{rank_stable}Dynamics of wed diffusion: stability of LAD}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{../../outputs/ranks/web_per_firm2000_2012_only95_av.png}

\caption{\label{rank_unstable}Dynamics of wed diffusion: volatility LAD}

Then, figures \ref{morani} and \ref{lisa} offer a first insight into
whether a neighbourhood effect underpins the diffusion of web
technologies in the UK as they plot the Moran's I and the LISA maps of
website density respectively for LAD and OA. Starting from the former,
spatial autocorrelation was higher in the beginning and then around year
2000 dropped slightly and stabled at 0.2. This reflects the early
concentration of high website density around London, which over time
diffused as high-high clusters can be seen in other parts of the country
away from London as seen in Figure \ref{lisa}. An almost reverse pattern
can be observed for OA. At the beginning of the study period Moran's I
was around 0.1 and it plateaued after 2000-2001 around 0.2. Because of
the very small size of OA, at the early stages of the diffusion of web
technologies their adoption was spatially scattered. This is reflected
in the lack of any significant clusters in 1996. Eventually, as the
adoption rate increased, more such clusters of high website density were
formed and this is reflected both in the Moran's I and the LISA maps in
Figures \ref{morani} and \ref{lisa}. All in all, the exploratory spatial
data analysis advocates towards an underpinning neighbourhood effect and
the different scales of analysis illustrate how it evolved differently
over time.

{\centering \includegraphics[width=1\textwidth,height=\textheight]{tranos2023_files/figure-pdf/morani-1.pdf}

\caption{\label{morani}Website density Moran's I}

Figure \ref{lisa}

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{tranos2023_files/figure-pdf/unnamed-chunk-7-1.pdf}

\caption{\label{lisa}\centering Website density LISA maps}

To illustrate whether a hierarchical process underpins the diffusion of
web technologies, the Gini coefficient is calculated yearly both for LAD
and OA. As a metric of inequality, the Gini coefficient demonstrates
whether website density is concentrated in a small number of LAD/OA, or
whether it is more equally spread across the country. Both scales of
analysis in Figure \ref{gini} illustrate the same picture. At the
beginning of the commercial internet website density was extremely
unequal, or in other words, only a few places had websites associated
with them. Inequality dropped and plateaued after 2000 for both scales
illustrating that at the first stages of the commercial internet,
website density was concentrated in a limited number of places. This is
an indication of a hierarchical diffusion mechanism that led over time
to a more equal spread. Interestingly, the year 2000 is again a period
of change for this diffusion mechanisms as it was for the neighbourhood
process. There is a substantial difference between the Gini coefficient
magnitude for LAD and OA, but this is well expected as the very small
size of OA equates to a lot polygons without any websites pointing to
them -- for instance residential OA. LAD/OA

{\centering \includegraphics[width=1\textwidth,height=\textheight]{tranos2023_files/figure-pdf/gini-1.pdf}

\caption{\label{gini}Website density Gini coefficient}

The next section incorporates the above discussed spatial processes of
the diffusion of web technologies into a modelling framework. The aim is
to use variables depicting these spatial processes in order to predict
the diffusion of web technologies in the UK over space and time and
across different scales. Specifically, four different models are
estimated. Firstly, all the data points for the OA and LAD are utilised
in order to build two respective RF models and assess their capacity to
predict the adoption of web technologies. These two models will reveal
the predictive capacity of the diffusion mechanisms and also allow to
see how the importance of such variables changes across scales. The next
two sets of models will be again trained on web diffusion at the two
working scales: OA and LAD. However, instead of using all the data
points, the OA and the LAD from one of the twelve UK regions are held
out and then the trained model is used to predict website density in the
OA or the LAD of the held-out region. This process takes place
recursively for all UK regions. The difference in the predictive
capacity of the different samples will reveal how dissimilar are these
spatial process across regions and, importantly, at different scales.

@@

The models include variables that capture the three processes that the
relevant literature and the descriptive analysis presented above
highlighted. Namely, the models capture: (i) a hierarchy effect with
diffusion running from main centres to secondary ones, (ii) a
neighborhood effect according to which diffusion first hits nearby
locations, and (iii) the rather canonical pattern of diffusion over time
as reflected in the S-shaped pattern in the cumulative level of
adoption.

To capture the hierarchy effect the models include as predictors a one
year lag of website density in London, the largest city in the UK, a one
year lag of the website density in the nearest city and the same for the
nearest retail centre. Due to the small sizes of the retail centres, the
latter is only relevant for the OA-level models. In addition, the models
include the distance to London, the nearest city and the nearest retail
centre. The underlying logic is that the level of website adoption in a
spatial unit depends on the level of the adoption in places further up
in the urban hierarchy the previous year. The inclusion of the distance
variables incorporates spatial structure into the hierarchy argument. To
depict the neighbourhood effect, the website density of the neighbouring
spatial units in the previous year is employed. Again, the underpinning
rationale is that the level of web adoption within a spatial unit
depends on the level of web adoption in the neighbouring spatial units
the year before. This represents the `hitting nearby locations first'
argument. Therefore, the spatial and temporal lag of the website density
in LAD and OA is calculated. Lastly, the time effect which is reflected
on the S-curve for the cumulative adoption is captured by a time trend
variable. Hence, all four model will follow the following generic form
(Eq. \ref{model}):
